Discussion Draft v1

### 2020-04-16 doc/Discussion.rmd
small edits

To place our results in the broader context of recombination rate variation we examine three modifier models and two functional and cell physiology models (table X). While there are challenges to fitting models built under different assumptions, we feel it's a useful exercise to highlight any unifying patterns for our specific traits across exsisting models. The three modifier models were built to explain variation in genome wide recombination rates between sexes (heterochiasmy), while the chromosome ocillatory movement (COM) model was built to describe positive interference. This spindle based selection model was developed to explain the results resented in this manuscript. In cases where the models can't be extended to predicting results, we list a 'NA' for the prediction.

@@

The direction of heterochiasmy under the two locus modifier model is dependent on female specific recombination modifier suppresses or increases recombination, if the modifier is linked to the driving locus, and the meiotic stage (reduction or equatorial division) when the driver acts (see Table 1 of @brandvain2012scrambling).

Under the spindle based model, faster male evolution of the genome wide recombination rate, is primarily driven by a more efficient spindle assembly checkpoint (SAC) acting on chromosome structures at metaphase I alignment. <We refer the reader to [@laneKauppi2019] for more details on the SAC pathway.> The SAC prevents anueploidy by ensuring all bivalents are correctly attached to the microtubulue, (bi-orientated) spindle before starting the metaphase-to-anaphase transition ([@laneKauppi2019], subramanian and Hochwagen , @dumontDesai2012).

A stable chromosome structure requires, i) kinetochores attached to opposite poles of the cells and ii) at least one crossover to create tension across a proportion of sister cohesion [@laneKauppi2019]. Unstable bivalents on the meiosis I spindle, cause the SAC to delay anaphase and allow time for the bivalent to stabilize on the microtubule spindle. If they fail to attach, the SAC triggers apoptosis. Genetic variants which increase the number of unstable bivalents (such as those affecting recombination rate or cohesion), will be selected against in favor for genetic variants causing faster stabilization of bivalents on the MI spindle. Multiple lines of evidence indicate that the effectiveness of SAC in spermatogenesis is greater compared to oogenesis ([@laneKauppi2019], Subrahman). These differences in SAC strength have been connected to conserved features of gametogenesis; centrosome spindle (@So2019, @dumontDesai2012) and cell volume [@kyogoku2017]. This model nominates conserved differences in in the context of meiosis between males and females as the primary source of distinct evolutionary trajectories in genome wide recombination rates.

@@

# consequence of greater female variance -- (between cell)
```

@@

As reviewed in [@sardell_sex_2020], sexual dimorphism in the broad scale recombination landscape is a highly conserved trait. The two locus modifier and SACE models predict sexual dimorphism in the broad scale recombination landscape, for diminishing the effect of drive systems and maintaining larger chromosome blocks under linkage in males respectively. The COM model, predicts the sex differences in the recombination landscape via interference is due to a combination of the length of the axis and differing strengths of chromosome movements during the pairing process [@hulten2011_COM].

Under the spindle based model we hypothesize that the sexual dimorphism in the recombination landscapes hinges on sex differences on the requirements for chromosome cohesion in late meiosis I. The irreversible process of the metaphase to anaphase transition is initiated by the protyltic decay of the sister cohesion connecting homologs [@laneKauppi2019], subramanian and Hochwagen , @dumontDesai2012]. The number and placement of crossovers alter the distribution of sister cohesion and the resulting chromosome structure when bivalents are aligned and bi-oriented on the metaphase plate [@vanVeen2003, @altendorfer2020]. Differences in timing of cell cycle between oogenesis and spermatogenesis imposes the the different selective pressures on how sister cohesion affects chromosome structure. Faster spermatogenesis may select for synchronization of the separation homologs (cite?). While in oogenesis, the slower cell cycle and multiple arrest stages may require chromosome structures with greater stability on the MI spindle, espicially for those with dicyate arrest [@lee2019].

@@

The COM model predicts that interference and the recombination landscape arises from known oscillatory movements during prophase, however it lacks a mechanism for a evolutionary based predictions. While the SACE modifier model does not explicitly model evolution of interference strength we note that a logical outcome of the main prediction of maintaining larger chromosome blocks in males, would be a recombination landscape with stronger interference strength.

We have focused on the model involving direct selection on the recombination pathway, which has an impact on the fitness of an individual, but we acknowledge connections to other models involving indirect selection on recombination rates which involve the fitness outcomes of offspring. A model from Goldstein et al, reviewed in [@ottoPaysuer2019] and @veller2019shuffle, predict that interference strength evolves whenever increased recombination rates evolve. The simulations under this model indicate that the space across multiple loci or between multiple crossovers increases in a positive manner with the genome wide recombination rate. A distinguishing feature of this model from other modifer models is that the number of crossovers per chromosome is kept constant. 
Empiracally, the range of crossovers per chromosome is quite limited, 1-3, for the majority of chromosomes across most taxa [@ottoPaysuer2019, @stapley_variation_2017]. Constraining the number of crossovers per chromosome in models may fit empirical data better than those where recombination rate across an abstracted genetic space is unconstrained.

@@

We make three suggestions for future steps in studying recombination variation. First consider the when comparing sex specific meiotic traits consider that evolutionary distinct trajectories due to the fundamental differences in gametogenesis. Second certain systems can be leveraged to overcome limitations inherent in some approached. For example, identifying chromosomes across cells in cytological data requires chromosome specific probes, (e.g. FISH) but chromosome specific data can be collected from organisms with diverse karyotypes (birds, peromyscus,  humans) or backgrounds with Robertsonian trans-locations.

>>>>

To place our results in the broader context of recombination rate evolution we compare our main resuts to expectionat under three modifier models and two functional and cell physiology models (table X). While there are challenges to fitting models built under different assumptions, we feel it's a useful exercise to highlight any unifying patterns for our specific traits across exsisting models. The three modifier models were built to explain variation in genome wide recombination rates between sexes (heterochiasmy), while the chromosome ocillatory movement (COM) model was built to describe positive interference. This spindle based selection model was developed to explain the results resented in this manuscript. In cases where the models can't be extended to predicting results, we list a 'NA' for the prediction.

@@

The direction of heterochiasmy under the two locus modifier model is dependent on female specific recombination modifier suppresses or increases recombination, if the modifier is linked to the driving locus, and the meiotic stage (reduction (MI) or equatorial division (MII)) when the driver acts (see Table 1 of @brandvain2012scrambling).

Under the spindle based model, faster male evolution of the genome wide recombination rate, is primarily driven by a more efficient spindle assembly checkpoint (SAC) acting on chromosome structures at metaphase I alignment. <We refer the reader to [@laneKauppi2019] for more details on the SAC pathway.> The SAC prevents anueploidy by ensuring all bivalents are correctly attached to the microtubulue spindle (bi-orientated) before starting the metaphase-to-anaphase transition [@laneKauppi2019, @subramanian2014, @dumontDesai2012].

A stable chromosome structure requires, i) kinetochores attached to opposite poles of the cells and ii) at least one crossover to create tension across a proportion of sister cohesion [@laneKauppi2019]. Unstable bivalents on the meiosis I spindle, cause the SAC to delay anaphase and allow time for the bivalent to stabilize on the microtubule spindle. If they fail to attach, the SAC triggers apoptosis. Genetic variants which increase the number of unstable bivalents (such as those affecting recombination rate or cohesion), will be selected against in favor for genetic variants causing faster stabilization of bivalents on the MI spindle. Multiple lines of evidence indicate that the effectiveness of SAC in spermatogenesis is greater compared to oogenesis [@laneKauppi2019], @subramanian2014]. These differences in SAC strength have been connected to conserved features of gametogenesis; presence of a centrosome spindle (@So2019, @dumontDesai2012) and cell volume [@kyogoku2017]. This model nominates conserved differences in in the context of meiosis between males and females as the primary source of distinct evolutionary trajectories in genome wide recombination rates.

@@

# consequence of greater female variance -- (between cell)

@@

As reviewed in [@sardell_sex_2020], sexual dimorphism in the broad scale recombination landscape is a conserved trait. The two locus modifier and SACE models predict sexual dimorphism in the broad scale recombination landscape, for diminishing the effect of drive systems and maintaining larger chromosome blocks under linkage in males respectively. The COM model, predicts the sex differences in the recombination landscape via interference is due to a combination of the length of the axis and differing strengths of chromosome movements during the pairing process [@hulten2011_COM].

Under the spindle based model we hypothesize that the sexual dimorphism in the recombination landscapes hinges on sex differences on the requirements for chromosome cohesion in late meiosis I. The irreversible process of the metaphase to anaphase transition is initiated by the protyltic decay of the sister cohesion connecting homologs [@laneKauppi2019], subramanian and Hochwagen , @dumontDesai2012]. The number and placement of crossovers alter the distribution of sister cohesion and the resulting chromosome structure when bivalents are aligned and bi-oriented on the metaphase plate [@vanVeen2003, @altendorfer2020]. Differences in timing of cell cycle between oogenesis and spermatogenesis imposes the the different selective pressures on how sister cohesion affects chromosome structure. Faster spermatogenesis may select for synchronization of the separation homologs (cite). While in oogenesis, the slower cell cycle and multiple arrest stages may require chromosome structures with greater stability on the MI spindle, espicially for those with dicyate arrest [@lee2019].

@@

The COM model predicts that interference and the recombination landscape arises from known oscillatory movements during prophase [@rubin2020mixing], however it lacks a mechanism for a evolutionary based predictions. While the SACE modifier model does not explicitly model evolution of interference strength we note that a logical outcome of the main prediction of maintaining larger chromosome blocks in males, would be a recombination landscape with stronger interference strength.

We have focused on the model involving direct selection on the recombination pathway, which has an impact on the fitness of an individual, but we acknowledge connections to other models involving indirect selection on recombination rates which involve the fitness outcomes of offspring. A model from @goldstein1993, reviewed in [@ottoPaysuer2019] and @veller2019shuffle, predict that interference strength evolves whenever increased recombination rates evolve. The simulations under this model indicate that the space across multiple loci or between multiple crossovers increases in a positive manner with the genome wide recombination rate. A distinguishing feature of this model from other modifer models is that the number of crossovers per chromosome is kept constant. 
Empiracally, the range of crossovers per chromosome is quite limited, 1-3, for the majority of chromosomes across most taxa [@ottoPaysuer2019, @stapley_variation_2017]. Constraining the number of crossovers per chromosome in models may fit empirical data better than those where recombination rate across an abstracted genetic space is unconstrained.

@@

We make three suggestions for future steps in studying recombination variation. First consider the when comparing sex specific meiotic traits consider that evolutionary distinct trajectories due to the fundamental differences in gametogenesis.

Second certain systems can be leveraged to overcome limitations inherent in some approached. For example, identifying chromosomes across cells in cytological data requires chromosome specific probes, (e.g. FISH) but chromosome specific data can be collected from organisms with diverse karyotypes (birds, peromyscus,  humans) or backgrounds with Robertsonian trans-locations.
